Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2024-09-10 Paper.qmd

xxxxxxxxxxDescribe camps brieflyxxxxxxxxxxxxxx - include dates and how we handled those

@@

Triceps and subscapular skinfold thicknesses were measured using skinfold calipers. Measurements were taken directly on the skin. The triceps measurement occurred at the midpoint of the upper arm. The subscapular measurement occured immediately below the inferior angle of the scapula, being picked up diagonally at the natural fold line of the skin [@antóna]. Flexed bicep circumference was measured perpendicular to the long axis of the upper arm. For this measure participants were asked to sit/stand? an position their upper arm perpendicular to their body with their elbow being help at a 90° angle, before flexing their clenched fist towards themselves as much as they could.

@@

When the cause involved punishment, however, parents were more likely to respond negatively and less likely to respond positively. In these circumstances, children may be signaling about the severity of a punishment [@syme2019]. However, parents may be reluctant to reduce punishments if the behavior that warranted the punishment was repeated or severe. They may also react with scolding or further punishment. That said, our model predicts positive responses in `r signif(out_mpr_full$estimate[3] * 100, 2)`% of cases (when all other predictors are held at their mean values), suggesting that children are successful in negotiating better outcomes at least some of the time.

@@

[\[MG1\]](#_msoanchor_1)Alternative start: Although exceptions exist (cite),

[\[MG2\]](#_msoanchor_2)Lumping together various models for the evolution of cooperation with either delayed benefits or inclusive fitness benefits. Not sure if this would confuse readers.

[\[MG3\]](#_msoanchor_3)Probably worth mentioning studies which fail to find evidence for it and why that might me (largely alloparenting benefits of other children)

[\[MG4\]](#_msoanchor_4)Find and cite others, if we decide this much detail is worth including. I am not sure it is.

[\[MG5\]](#_msoanchor_5)Removed paragraph over conflict over marriage partners due to it not seeming highly relevant given our data.

[\[MG6\]](#_msoanchor_6)Probably worth adding (see xxxxx for exceptions)

[\[MG7\]](#_msoanchor_7)Might also be worth mentioning signals between siblings (e.g., Roulin 2001 or the negotiating need section in Wells 2003).

[\[MG8\]](#_msoanchor_8)Consider reordering so that the adaptive problem precedes crying?

[\[MG9\]](#_msoanchor_9)Not sure if this paragraph is worth the space.

[\[MG10\]](#_msoanchor_10)Hadza children appear to be an exception  (Blurton Jones paper)

[\[MG11\]](#_msoanchor_11)Comment in Soltis. Also worth fitting in the possibility that children may use crying to annoy parents until they give in.

[\[MG12\]](#_msoanchor_12)Del Giudice, 2018 too?

# Cut from paper:

## Intro

### Conflicts of interest within the family

Conflicts of interest between children and parents may also involve how much children invest in the household. As previously mentioned, each individual child is expected to prefer greater investment in oneself than in one’s siblings, while parents, who are equally related to all their children, should often be expected to favor a more even distribution [@trivers1974]. This can lead to conflict with older children who might prefer to spend their time in ways which are likely to result in future direct fitness benefits (e.g., finding romantic partners or forming strong bonds with friends) rather than taking care of their siblings or acquiring or preparing food. Such a dynamic has been found among the Chuukese, with both male and female children reporting conflict with their families over children’s desires to invest financial resources in their own status enhancing opportunities rather than in the household [@lowe2003]. Similarly, Syme and Hagen [@syme2023a] found greater numbers of siblings were associated with increased likelihoods of severe conflict in a retrospective study with Chuukese-Americans. This may also be consistent with parents pressuring older children to invest in the household; however, there are other ways in which limited resources can result in behavioral conflict [@syme2023a].

Conflicts of interest over investment in the household also appear to extend to when a child starts having their own children [@kushnick2023; @emlen1995]. Due to the benefits children may provide their siblings, parents may be motivated to delay or subvert the reproductive careers of certain offspring [@flinn1989; @emlen1992], a conflict which is likely enhanced in families with step-parents due to the likelihood of future children being step-siblings rather than full siblings, thereby reducing the potential fitness benefits of alloparenting [@moya2014; @schaffnit2021]. Situations can also arise in which it might pay parents for their children to start having their own children sooner than would be ideal for the adult child. For example, parents may be motivated for a child to leave the household and begin having children earlier than is ideal from the child’s perspective if their presence in the household leads to them taking investment from other children who caretakers deem more worthy of investment [@kushnick2023].

Studies testing the existence of familial conflict over the timing of child reproduction are rare and direct tests on the existence of parent-offspring conflict over when a child should start their own reproductive career versus help existing siblings in humans are needed [@schaffnit2021]. The bulk of evidence for the existence of such conflicts comes from work with non-human animals. For example, the cooperative breeding white-fronted bee-eaters (*Merops bullockoides*) will interrupt the courtship of their offspring, seemingly to continue to benefit from their undiluted assistance in digging nests, incubating the clutch, finding food for nestlings, and defending the nest [@emlen1992]. When it comes to humans, Flinn found that rural Trinidadian females contributed substantial amounts of childcare and those with siblings under 10 years old had their first child at a later age [@flinn1989], both of which would be expected if there is a conflict of interest over investing in one’s siblings or one’s own reproduction. However, a study among Karo Batak women found no evidence of delayed reproduction for alloparenting benefits [@kushnick2006].

One possibility is that crying is physiologically constrained in ways that prevent or limit faking. Evidence consistent with this comes from findings that crying is hard to do on command and findings individuals often cry while alone, something which is not predicted by signaling theory [@vingerhoets2001]. However, the likelihood of crying is often clearly related to presence of supporters, something which is seen in findings that lonely individuals cry less than those with more support options despite reporting lower well-being [@vingerhoets2013] and the existence of delayed crying—times when crying is inhibited during the cause of negative affect only for it to be expressed when one is near potential supporters [@gracanin2018]. A more likely possibility is that certain acoustic elements of infant cries are tied to their physiological state in ways that make faking certain states impossible, with crying serving as a graded signal in most other instances [@soltis2004; @zeifman2017].

It is likely that the functions of crying and the mechanisms behind its ability to elicit pro-signaler responses change over one’s life. Compared to adults and older children, the physiological costs of crying are greater for newborns and infants, among whom crying can result in a substantial increase in energy expenditure [@rao1997; @thureen1998]. Due to these costs, crying at certain intensities or for long periods of time may be prohibitively costly for infants with low survival prospects, making it a potential credible signal of vigor [@furlow1997; @lummaa1998; @soltis2004; @wells2003]. Such signals are thought to be important due to parental adaptations which are suspected to take the viability of offspring into account when making decisions about whether it is worth investing in a specific child or withdrawing investment to instead invest in other children who are more likely to survive to reproductive age and be more competitive adults [@hagen1999; @lummaa1998]. In line with this reasoning, most infanticides cross-culturally occur when a child has poor survival prospects, due to either health conditions or resource constraints [@daly1984; @soltis2004].

Another often suggested means by which crying could result in credible signals of need is if the increased risk of predation that comes from loud vocalizations allows for signaling to be worthwhile for those in relatively high need but not for those with low need [@lummaa1998; @wells2003; @zahavi1977]. Evidence of increased predation has been found in studies on birds [@leech1997; @lima2009; @redondo1992; @ryan1982], among whom louder broods in general have also been found to be more conspicuous to predators [@haskell1997; @leech1997]. It is unclear how significant the risk of alerting predators would be ancestrally. However, there are ethnographic reports of mothers keeping infants and young children close in dangerous environments [@hewlett1998; @kaplan1987; @konner1972], something which likely lessens this risk.

As children age, their reproductive value increases due to reduced odds of mortality, leading to a decreased risk of infanticide (Hagen 2004). For this reason, there is likely a transition period in which crying begins to convey information about how much a child values further investment. Many models have been put forward which attempt to explain human crying and begging in other species as credible signals of need due to the differential costs and benefits of crying and analogous vocalizations in other species as deterring the exaggeration of need [@godfray1991; @kilner1997; @parker2002]. Although differing in their details, a common theme is that the indirect fitness costs signalers incur from imposing costs on that parents and siblings through begging (be them energetic costs, survival costs, or reduced likelihood of future siblings) creates a floor to possible deception but falls short of preventing it [@daly1990; @wells2003].

When it comes to older children, the relatively insignificant energetic costs of crying are unlikely to be enough to result in credible signals in times of substantial conflicts of interest. For this reason, crying among older children may be more similar to adult crying, something which has been found to result in increased support and empathy but sometimes come at the cost of reputation and the potential for stigmatization [@cornelius2001; @cretser1982; @hendriks2006; @jesser1989; @vingerhoets2016]. Such costs may be a means by which crying can still be a credible signal [@sznycer] with moderate conflicts of interest if they dissuade potential fakers from making the signal.

## Discussion

### Running away

Probably cut When viewed in isolation, the frequencies of sadness, crying, and temper tantrums were also positively associated with conflict frequency between child and caretaker. However, the frequency of running away was not significantly associated with conflicts with parents. A retrospective study of parent-offspring conflict among Chon Chuuk immigrants in the northwest United States found that running away was the most common response to long-standing conflicts with parents and often resulted in beneficial outcomes for the child [@syme2023a]. The threat running away poses to child wellbeing and, therefore, its ability to credibly signal information about one’s current situation is likely highly dependent on one’s socioecology and age. Among the Chon Chuuk, running away is often seen as a suicide threat [@fischer1950; @syme2023a], likely due to xxxxxxxxx. Similar risks may also occur in environments with large predators, venomous animals, or high degrees of human conflict. In these instances, even one night away can involve substantial risk as is seen among the !Kung where a woman running away and spending the night outside of camp can be used to communicate “the strength of her feelings” about a an unhappy relationship [@shostak1981]\[\[6\]\]. When it comes to Utila, mothers in focus groups reported that running away on Utila was much less risky than it would be on mainland Honduras, with a few mothers mentioning that they warned their children about the difference in danger that came with running away on the mainland. Despite the relative safety of Utila, young children may still face risks out on their own due to proximity to the ocean, large dogs, and fast-moving motorcycles, golf carts, and tuk-tuks on the main road. For these reasons, running away may be too costly for young children but not costly enough to credibly signal need for older children.

### Perceived immune function

A negative association between caretaker estimates of child immune function and child signaling was seen, suggesting that signaling behavior was influenced by adversity. Perceptions of greater immune susceptibility were associated with both the summed costliness of signaling and the frequency of the most common child signal of need, while the frequency of child signaling was marginally significant. When examining the effect of perceived immune function on each signal, perceptions of greater immune susceptibility predicted the frequency of sadness but not the frequencies of crying and temper tantrums.

One explanation for this may be that parents can easily detect child illness and that conflicts of interest tend to be less significant when it comes to illness care than other areas. Pathogens have been a strong selection pressure for life on earth for much of its existence and have greatly shaped evolutionary history (cite). In response to this, humans have evolved many adaptations for limiting the harm we experience from pathogens, including the ability to identify relaible indicators of illness in others. In some contexts, this activates disgust and motivates avoidance of sick individuals. However, it can also lead to increased care of relatives and allies (cite), with there being some suggestion that we have even evolved to upregulate some of our symptoms beyond what would arise from an immune response/pathogen interaction to better convince others that we are sick (cite). Due to this evolutionary history, parents may be better equipped to identify the possible benefits to investing in a sick child when compared to investment aimed at helping a child in status competition (e.g., buying new or popular clothing), something which may be exacerbated by culture change in what is popular among children. In other words, there is less private information for a child to reveal. This coupled with the likelihood that conflicts of interest between children and caretakers are smaller when it comes to illness care compared to other areas means there is less to gain from costlier signals.

### Cross-sectional study

Another limitation is that our study was cross-sectional. A major challenge when studying signals of need is the fact that such signaling is expected to 1) be more common in times when individuals can benefit from increased investment (e.g., when they are facing major adversity or when a fitness-relevant opportunity arises) and 2) result in benefits to the signaler on average. In other words, higher levels of signaling may stem from greater levels of need while simultaneously eliciting more investment from others. This hypothesized causal, bi-directional relationship makes it difficult to isolate the effects of signaling behavior on investment in children when using standard measures of child welfare (i.e., BMI, health metrics, and height for age). Longitudinal studies can help address this. However, our small sample size of participants with signaling data and anthropometric data at two time periods prevents this for this study.

>>>>

xxxxxxxxxxDescribe camps brieflyxxxxxxxxxxxxxx - include dates and how we handled those. For cross-sectional analyses, we used the 2023 anthropometric values due to them being closer in time to the caretaker surveys.

@@

Triceps and subscapular skinfold thicknesses were measured using skinfold calipers. Measurements were taken directly on the skin. The triceps measurement was taken at the midpoint of the upper arm. The subscapular measurement occurred immediately below the inferior angle of the scapula, being picked up diagonally at the natural fold line of the skin [@antóna]. Flexed bicep circumference was measured perpendicular to the long axis of the upper arm. For this measure participants were asked to sit/stand? an position their upper arm perpendicular to their body with their elbow being help at a 90° angle, before flexing their clenched fist towards themselves as much as they could.

@@

When the cause involved punishment, however, parents were more likely to respond negatively and less likely to respond positively. In these circumstances, children may be signaling about the severity of a punishment [@syme2019]. However, parents may be reluctant to reduce punishments if the behavior that warranted the punishment was repeated or severe. They may also react with scolding or further punishment. That said, our model predicts positive responses in `r signif(updated_out$estimate[3] * 100, 2)`% of cases (when all other predictors are held at their mean values), suggesting that children are successful in negotiating better outcomes at least some of the time.

@@
